---
title: "MENASA Risk Monitor — Independent Validation in R"
author: "Muhammad Rafay Waqar"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

The MENASA Risk Monitor's own in-app Methodology tab runs two validation checks on
its 11-factor composite score, both written in Python: a principal component
analysis (PCA) redundancy check across the 11 factors, and a historical backtest
against six well-documented sovereign crises. This notebook reproduces both checks
**independently, in R**, against the exact same underlying data files
(`driver_history.csv`, `scored_history.csv`) committed to this repository —
not a re-explanation of the Python result, a second, separately-coded
computation of it.

Where the two disagree, that would be worth knowing. Where they agree, that's a
real (if modest) form of external validation: the same finding, reached by an
independently written implementation in a different language and statistical
package (R's `plm` and base `prcomp`, rather than the app's hand-rolled
`numpy`/`pandas` eigendecomposition).

```{r libs}
library(plm)
data_dir <- "../"
```

## 1. Independent PCA Redundancy Check

The live app computes this via a manual eigendecomposition of the standardized
factor correlation matrix (`numpy.linalg.eigh`), complete-case only. Here, the
same complete-case data is fed into R's own `prcomp()`, a completely separate
implementation.

```{r pca}
factor_cols <- c(
  "debt_to_gdp", "current_account_pct_gdp", "reserves_months_imports",
  "gdp_growth", "inflation", "currency_depreciation_pct",
  "political_stability", "government_effectiveness", "rule_of_law",
  "regulatory_quality", "control_of_corruption"
)

driver_history <- read.csv(file.path(data_dir, "driver_history.csv"))
complete <- driver_history[complete.cases(driver_history[, factor_cols]), factor_cols]

cat("Complete-case observations used:", nrow(complete), "of", nrow(driver_history), "\n")

pca <- prcomp(complete, scale. = TRUE)
var_explained <- (pca$sdev^2) / sum(pca$sdev^2) * 100

cat("\nVariance explained by each principal component (%):\n")
print(round(var_explained, 1))

cat("\nPC1 loadings (which factors move together on the dominant component):\n")
pc1 <- sort(pca$rotation[, 1], decreasing = TRUE)
print(round(pc1, 3))
```

```{r pca-plot, fig.width=7, fig.height=4}
barplot(var_explained, names.arg = paste0("PC", seq_along(var_explained)),
        col = "#0d1b2a", border = NA, ylab = "Variance Explained (%)",
        main = "Independent R Reproduction: Variance Explained by Component")
```

```{r pca-corr}
cat("\nGovernance-cluster correlations (the app's claimed redundancy):\n")
cat("Rule of Law vs. Control of Corruption:  ",
    round(cor(complete$rule_of_law, complete$control_of_corruption), 2), "\n")
cat("Rule of Law vs. Political Stability:    ",
    round(cor(complete$rule_of_law, complete$political_stability), 2), "\n")
cat("Govt. Effectiveness vs. Control of Corr:",
    round(cor(complete$government_effectiveness, complete$control_of_corruption), 2), "\n")
```

**Read against the app's stated result:** the live Methodology tab reports PC1
explaining a dominant share of variance, driven by the four governance factors
(Rule of Law, Control of Corruption, Political Stability, Government
Effectiveness) loading together with similar strength (0.38–0.45 each), while
the economic factors split onto later components. The numbers above, computed
independently in R, are the same check — compare the printed values directly
against the live app's Methodology tab at the time this notebook was run.

## 2. Independent Historical Crisis Backtest

The app claims: of six well-documented crises, four were captured in the same
calendar year the crisis began (Syria 2011→12, Sri Lanka 2022, Afghanistan
2021, Pakistan 2022), one with roughly a one-year lag (Lebanon), and one still
missed (Egypt). This section recomputes each year-over-year delta directly
from `scored_history.csv`, independently of the app's own code.

```{r backtest}
scored_history <- read.csv(file.path(data_dir, "scored_history.csv"))

crisis_cases <- data.frame(
  code = c("SYR", "LKA", "AFG", "LBN", "PAK", "EGY"),
  crisis_year = c(2012, 2022, 2021, 2020, 2022, 2022),
  label = c("Syria (civil war escalation)", "Sri Lanka (2022 collapse & default)",
            "Afghanistan (2021 Taliban takeover)", "Lebanon (2019-23 financial collapse)",
            "Pakistan (2022-23 BoP crisis)", "Egypt (2022-23 currency crisis)")
)

for (i in seq_len(nrow(crisis_cases))) {
  code <- crisis_cases$code[i]
  yr <- crisis_cases$crisis_year[i]
  ch <- scored_history[scored_history$country_code == code, c("year", "risk_score")]
  ch <- ch[order(ch$year), ]
  prior <- ch$risk_score[ch$year == (yr - 1)]
  curr  <- ch$risk_score[ch$year == yr]
  delta <- round(curr - prior, 1)
  cat(sprintf("%-45s %d -> %d: %+.1f points\n", crisis_cases$label[i], yr - 1, yr, delta))
}
```

**Read against the app's stated result:** the deltas above are computed fresh
from the committed CSV, independently of the Streamlit app's own rendering
logic. They should match the app's stated per-case figures (e.g., Sri Lanka
+13.9 in 2022, Syria +22.0 in 2012) exactly, since both draw on the same
underlying `scored_history.csv` — the value of this check is confirming no
transcription or rendering error crept into how the app displays its own
backtest, not discovering a new result.

## 3. Panel Structure Check (`plm`)

As a genuinely different addition (not just a re-run of the app's own checks),
this confirms the data is a valid balanced/unbalanced panel and reports basic
panel diagnostics using R's `plm` package — the standard tool for exactly this
kind of country-year data, and the tool the planned EM Macro & Geopolitical
Risk Engine's sovereign distress model will build on directly.

```{r panel}
pdata <- pdata.frame(scored_history, index = c("country_code", "year"))
cat("Panel dimensions:\n")
print(pdim(pdata))
```

## Limitations of This Notebook

- This is a validation and cross-check exercise, not a new statistical claim.
  It confirms the app's own reported numbers reproduce under an independently
  written implementation; it does not assess whether the underlying 11-factor
  weighting scheme itself is the right one (the app's own README is explicit
  that it isn't econometrically validated).
- The panel check above is illustrative of the data's structure, not a fitted
  model — the actual sovereign distress model (fixed/random-effects regression
  with real out-of-sample validation) is a separate, larger project.
